---
title: "ggplot2_tutorial"
author: "Adam Chmurzynski"
date: "December 6, 2017"
output: html_document
---

**Lesson on ggplot2**
by Adam Chmurzynski

In this exercise we will work with a dataset iris that is available in R. Follow the steps below to upload the appropriate library and datasets.

Plots take data frames as input, allowing us to visualize columns and rows. There are two main elements: aethetics(aes) and geoms
Use the '+' operator (think of adding elements to your figure) to 
concatenate plots, geoms and scales into complex images.


```{r}
library(ggplot2)
data(iris)
head(iris)

iris_plot_0 <- ggplot(iris, aes(x=Sepal.Length, y=Petal.Length)) + 
                geom_point(size = 1) + geom_line()
```

Geoms are visual elements we use to look at rows (observations)
Aesthetics are visual features used to manipulate geoms by column

```{r}
iris_plot_1 <- ggplot(iris, aes(x=Sepal.Length, y=Petal.Length)) + 
  geom_point(aes(shape=Location)) +
  geom_point(aes(color=Species, size=Petal.Width))

mm <- lm(Petal.Length~Sepal.Length, iris)

iris_plot_2 <- ggplot(iris, aes(x=Sepal.Length, y=Petal.Length)) + 
  geom_point(aes(color=Species, size=Petal.Width)) +
  geom_smooth(method="lm", se=FALSE, color="black")

```
Scales are applied to aesthetics to map numerical ranges from data to styles 

```{r}
iris_plot_3 <- ggplot(iris, aes(x=Sepal.Length, y=Petal.Length)) + 
  geom_point(size = 2, aes(color=Sepal.Length, alpha=Sepal.Width)) +
  scale_x_continuous(limits=c(0, 10)) +
  scale_y_continuous(limits=c(0, 5)) +
  scale_color_continuous(low="#728D23", high="#AB74D0") 
  #scale_alpha_continuous(range=c(0.05, 0.5))
```

Faceting creates multiple plots of pairwise comparisons along a factor
```{r}
iris_plot_0 <- iris_plot_0 + facet_grid(Species ~ .)
iris_plot_4 <- iris_plot_0 + 
  #geom_text(p-values, aes(Species))
  facet_grid(Species ~ .) +
  #geom_smooth(method="loess", se=TRUE, color="red", fullrange=TRUE)
  geom_smooth(method="lm", se=TRUE, color="blue", fullrange=TRUE)
  
print(iris_plot_0)
  
```







