---
title: "Structural equation modeling"
author: "Sandra M Durán"
date: "December 22, 2017"
output: html_document
---
**Lesson on structural equation modeling (SEM) in R- Dec 7 2017**

In this script, I cover some basics of SEM in R, specifically SEM with observed variables. The script and slides focus on developing models, and understanding model construction. 

This script was developed using some online tutorials and R-code available from Dr. Jarret Byrnes and Bill Shipley. To get help in other types of SEM, and use of other packages (e.g., lavaan) check the references and links below:

+ [Jarret Byrnes tutorials](http://jarrettbyrnes.info/sem.shtml)
+ [Shipley 2009. Confirmatory path analysis paper](http://onlinelibrary.wiley.com/doi/10.1890/08-1034.1/abstract) Usefult to perform path analysis with categorical or presence/absence data)
+ Shipley, B. 2016. Cause and correlation in biology: a user's guide to path analysis, structural equations and causal inference with R. Cambridge University Press.
+ Grace, James B. Structural equation modeling and natural systems. Cambridge University Press, 2006.
+ [Piecewise SEM package](https://jonlefcheck.net/2014/07/06/piecewise-structural-equation-modeling-in-ecological-research/) Useful to learn model specification and run SEMs with no-continuous data. 

***
**Basic definitions of SEM**

Review attached slides 1-5 for basic definitions of SEM


```{r cars}
summary(cars)

#Set your working directory
setwd("C:/Users/smduranm/Documents/R-lessons/EEB_R_workshops/data")


#uploading required packages
library(lavaan)
library(car)
library(QuantPsyc)
library(ggm)
library(semPlot)
library(devtools)

#Load your data file
keeley<-read.csv("./Keeley_rawdata.csv")

#check the data
head(keeley)
summary(keeley) #missing data?

#Analyze the relationship between variables
pairs(keeley) #linear relationships, normal dist, outliers
myvars <- keeley[, c(4,6,7)]
pairs(myvars)

#Our model for today
names(keeley)
attach(keeley)

#regression syntax versus lavaan
aLM<-lm(cover ~ age, data=keeley)

#F Table
library(car)
Anova(aLM)

#getting coefficients
summary(aLM)

#standardized coefficients
coef(aLM)[2]*sd(keeley$age)/sd(keeley$cover)

#standardized coefficients 2
library(QuantPsyc)
lm.beta(aLM)


#####
#Multiple Regression & SEM
#####

aLM2<-lm(cover ~ age+firesev, data=keeley)

#what to include...
Anova(aLM2)

#difference with anova!
anova(aLM2)

#standardized coefs
cor(keeley$age, keeley$firesev)
summary(aLM2)$r.squared
lm.beta(aLM2)

#finish the sem
aLM3<-lm(firesev ~ age, data=keeley)
Anova(aLM3)
summary(aLM3)$r.squared
lm.beta(aLM3)

##############################
#Moving to SEM
##############################

#Refit the new cover relationship
fullMedLM<-lm(cover ~ firesev, data=keeley)
summary(fullMedLM)$r.squared
lm.beta(fullMedLM)

keeley$fireCoverResiduals<-residuals(fullMedLM)
residLM<-lm(fireCoverResiduals ~ age, data=keeley)
Anova(residLM)
plot(fireCoverResiduals ~ age, data=keeley)


################################
### SEM for Exercise
################################

#fit the pieces
abioticLM <- lm(abiotic ~ distance, data=keeley)
heteroLM <- lm(hetero ~ distance, data=keeley)
richnessLM <- lm(rich ~ abiotic + distance + hetero, data=keeley)

#evaluate the pieces
Anova(abioticLM)
Anova(heteroLM)
Anova(richnessLM)

#standardized coefficients & r^2
lm.beta(abioticLM)
lm.beta(heteroLM)
lm.beta(richnessLM)

summary(abioticLM)$r.squared
summary(heteroLM)$r.squared
summary(richnessLM)$r.squared

#####
#Evaluate Mediation
richnessLM2 <- lm(rich ~ abiotic  + hetero, data=keeley)
summary(richnessLM2)$r.squared
lm.beta(richnessLM2)

keeley$richnessResidual<-residuals(richnessLM2)
richResidLM<-lm(richnessResidual ~ distance, data=keeley)
Anova(richResidLM)

#plot result
plot(richnessResidual ~ distance, data=keeley, cex=1.5)
abline(richResidLM, col="red", lwd=2)










```


```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
